---
editor_options:
  markdown:
    wrap: 72
---

# Practical 3: The Quality of Red Bordeaux Vintages {.unnumbered}

*This practical was developed by Dr. Tatjana Kecojevic, Lecturer in Social Statistics.*

We would like to analyse the quality of a vintage that has been quantified as the price and make the interpretation of our statistical findings. 

We are going to use `wine.csv` available from Eduardo García Portugués’s book: *Notes for Predictive Modelling; https://bookdown.org/egarpor/PM-UC3M/*. 

The **wine** dataset is formed by the auction price of 27 red Bordeaux vintages, five vintage descriptors (WinterRain, AGST, HarvestRain, Age, Year), and the population of France in the year of the vintage, FrancePop.
  
- Year: year in which grapes were harvested to make wine   
- Price: logarithm of the average market price for Bordeaux vintages according to 1990–1991 auctions. The price is relative to the price of the 1961 vintage, regarded as the best one ever recorded  
- WinterRain: winter rainfall (in mm)  
- AGST: Average Growing Season Temperature (in degrees Celsius)  
- HarvestRain: harvest rainfall (in mm)  
- Age: age of the wine measured as the number of years stored in a cask   
- FrancePop: population of France at Year (in thousands)  

You will require the `GGally` package; please make sure to install it first. 

```{r warning = FALSE, message = FALSE}
library(GGally)
```

And now let's import the data.

```{r}
wine <- read.csv("https://raw.githubusercontent.com/egarpor/handy/master/datasets/wine.csv")
```

Let's first obtain some summary statistics.  

```{r}
summary(wine)
```

And a matrix of plots. The `ggpairs()` function which is part of the `GGally` package produces an information rich visualisation that includes pairwise relationships of all the variables in the dataset.

```{r}
ggpairs(wine)
```


:::question
What conclusions can you draw based on the above information?
:::


We are going to investigate possible interactions between the rainfall (WinterRain) and the growing season temperature (AGST). We will start with the most complicated model that includes the highest-order interaction. In R we will specify the three-way interaction, which will automatically add all combinations of two-way interactions.  


```{r}
model1 <- lm(Price ~ WinterRain + AGST + HarvestRain + Age + WinterRain * AGST * HarvestRain, data = wine)

summary(model1)
```

:::question
What can you say about the explained variability of the model? Which coefficients are statistically significant? Simplify the model in stages and decide on the final model. 
:::

