---
editor_options: 
  markdown: 
    wrap: 72
---

# (PART\*) Section 1 {.unnumbered}

# Overview {.unnumbered}

::: {style="color: #333; font-size: 24px; font-style: italic; text-align: justify;"}
Section 1: Introduction to Data Science: The Basics of Statistical
Learning
:::

**In this section, we will cover the following functions:**

|   Function    |               Description               | Package |
|:-------------:|:---------------------------------------:|:-------:|
|     `c()`     |               concatenate               | base R  |
|    `ls()`     | provides names of object in environment | base R  |
|    `rm()`     |             remove objects              | base R  |
|   `View()`    | viewing data objects spreadsheet-style  | base R  |
| `read_spss()` |         reading SPSS data files         | `haven` |
| `read_dta()`  |        reading Stata data files         | `haven` |
| `read.csv()`  |          reading `.csv` files           | base R  |
|  `readRDS()`  |          reading `.RDS` files           | base R  |
|   `load()`    |         loading `.RData` files          | base R  |
|   `save()`    | saving multiple files (`.Rdata` format) | base R  |
|  `saveRDS()`  |   saving single files (`.RDS` format)   | base R  |
| `write.csv()` |       save a file as `.csv` file        | base R  |

::: ilos
**Learning Outcomes:**

-   Employ the basics of the R language to run simple lines of code;
-   Import different types of data files;
-   Export data in different formats.
:::

# Practical {-}

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var answers = document.querySelectorAll('.answers');

    answers.forEach(function(answer) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-answer-button';
        button.textContent = 'Show Answers'; // Updated text content
        button.style.display = 'block'; // Ensure button is visible
        answer.style.display = 'none'; // Initially hide the answer

        // Add click event listener to the button
        button.onclick = function() {
            if (answer.style.display === 'none') {
                answer.style.display = 'block'; // Show the answer
                button.textContent = 'Hide Answers'; // Update button text
            } else {
                answer.style.display = 'none'; // Hide the answer
                button.textContent = 'Show Answers'; // Reset button text
            }
        };

        // Insert the button before the answer
        answer.parentNode.insertBefore(button, answer);
    });
});

```

For this practical, you will use the `college.csv` file to refresh your practical programming skills. This file contains 18 variables for 777 different universities and colleges in the US.  

The variables are:  

-   Private : Public/private indicator
-   Apps : Number of applications received
-   Accept : Number of applicants accepted 
-   Enroll : Number of new students enrolled  
-   Top10perc : New students from top 10% of high school class
-   Top25perc : New students from top 25% of high school class
-   F.Undergrad : Number of full-time undergraduates
-   P.Undergrad : Number of part-time undergraduates
-   Outstate : Out-of-state tuition
-   Room.Board : Room and board costs
-   Books : Estimated book costs
-   Personal : Estimated personal spending
-   PhD : Percent of faculty with Ph.D.’s
-   Terminal : Percent of faculty with terminal degree
-   S.F.Ratio: Student/faculty ratio
-   perc.alumni : Percent of alumni who donate
-   Expend : Instructional expenditure per student
-   Grad.Rate : Graduation rate


```{r college-solution}
#a
college <- read.csv("https://www.statlearning.com/s/College.csv")
#b
rownames(college) <- college[, 1]
college <- college[, -1]

head(college)

#c
#i)
summary(college)

#ii)
pairs(college[,3:12])

# because the first two columns are not numerical vectors
#Please carefully check the data when you have similar errors

#iii) plot side by side boxplots of Outstate versus Private (Private universities have more out of state students):

class(college$Private)
#Character vectors cannot be used in most statistical functions in R, since they don't define groups.
college$Private <- ifelse(college$Private=="Yes", 1, 0)
#we need to convert the Character vector 'private' to numerical vector.
plot(college$Private, college$Outstate)

#iv)
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)

summary(Elite)

# plot side by side boxplots of Outstate versus Elite (Elite universities have more out of state students):
plot(college$Elite, college$Outstate)

#v)

par(mfrow=c(2,2))
hist(college$Apps)
hist(college$perc.alumni, col=2)
hist(college$S.F.Ratio, col=3, breaks=10)
hist(college$Expend, breaks=100)

#vi)
#for example

college[which.max(college$Top10perc), ]  # what is the university with the most students in the top 10% of class


acceptance_rate <- college$Accept/college$Apps

college[which.min(acceptance_rate), ]  # what university has the smallest acceptance rate

college[which.max(acceptance_rate), ]  # what university has the most liberal acceptance rate

```

## 2.4 Q10

This exercise involves the Boston housing data set.

```{r boston, exercise=TRUE}
#Q(a)-(d)

```

```{r boston-solution}
#a
library(MASS)
?Boston
dim(Boston)

#b
pairs(Boston)

#c
plot(Boston$age, Boston$crim)
# Older homes, more crime
plot(Boston$dis, Boston$crim)
# Closer to work-area, more crime
plot(Boston$rad, Boston$crim)
# Higher index of accessibility to radial highways, more crime
plot(Boston$tax, Boston$crim)
# Higher tax rate, more crime
plot(Boston$ptratio, Boston$crim)

#d
par(mfrow=c(1,3))
hist(Boston$crim[Boston$crim>1], breaks=25)
# most cities have low crime rates, but there is a long tail: 18 suburbs appear
# to have a crime rate > 20, reaching to above 80
hist(Boston$tax, breaks=25)
# there is a large divide between suburbs with low tax rates and a peak at 660-680
hist(Boston$ptratio, breaks=25)
# a skew towards high ratios, but no particularly high ratios

```

```{r boston2, exercise=TRUE}
#Q(e)-(h)


```

```{r boston2-solution}
# (e)
dim(subset(Boston, chas == 1))

# (f)
median(Boston$ptratio)

# (g)
t(subset(Boston, medv == min(Boston$medv)))
summary(Boston)
# Not the best place to live, but certainly not the worst.

# (h)
dim(subset(Boston, rm > 7))
dim(subset(Boston, rm > 8))
summary(subset(Boston, rm > 8))
summary(Boston)
# relatively lower crime (comparing range), lower lstat (comparing range)

```
