---
editor_options:
  markdown:
    wrap: 72
---


In Chapter 4, we used logistic regression to predict the probability of default using income and balance on the Default data set. We will now estimate the test error of this logistic regression model using the validation set approach. Do not forget to set a random seed before beginning your analysis.

Fit a logistic regression model that uses income and balance to predict default.

```{r}
library(ISLR2)
set.seed(42)
fit <- glm(default ~ income + balance, data = Default, family = "binomial")
```

Using the validation set approach, estimate the test error of this model. In order to do this, you must perform the following steps:
Split the sample set into a training set and a validation set.
Fit a multiple logistic regression model using only the training observations.
Obtain a prediction of default status for each individual in the validation set by computing the posterior probability of default for that individual, and classifying the individual to the default category if the posterior probability is greater than 0.5.
Compute the validation set error, which is the fraction of the observations in the validation set that are misclassified

```{r}
train <- sample(nrow(Default), nrow(Default) / 2)
fit <- glm(default ~ income + balance, data = Default, family = "binomial", subset = train)
pred <- ifelse(predict(fit, newdata = Default[-train, ], type = "response") > 0.5, "Yes", "No")
table(pred, Default$default[-train])
```

```{r}
mean(pred != Default$default[-train])
```

Repeat the process in (b) three times, using three different splits of the observations into a training set and a validation set. Comment on the results obtained.

```{r}
replicate(3, {
  train <- sample(nrow(Default), nrow(Default) / 2)
  fit <- glm(default ~ income + balance, data = Default, family = "binomial", subset = train)
  pred <- ifelse(predict(fit, newdata = Default[-train, ], type = "response") > 0.5, "Yes", "No")
  mean(pred != Default$default[-train])
})
```

The results obtained are variable and depend on the samples allocated to training vs. test.

Now consider a logistic regression model that predicts the probability of default using income, balance, and a dummy variable for student. Estimate the test error for this model using the validation set approach. Comment on whether or not including a dummy variable for student leads to a reduction in the test error rate.

```{r}
replicate(3, {
  train <- sample(nrow(Default), nrow(Default) / 2)
  fit <- glm(default ~ income + balance + student, data = Default, family = "binomial", subset = train)
  pred <- ifelse(predict(fit, newdata = Default[-train, ], type = "response") > 0.5, "Yes", "No")
  mean(pred != Default$default[-train])
})
```

Including student does not seem to make a substantial improvement to the test error.



We continue to consider the use of a logistic regression model to predict the probability of default using income and balance on the Default data set. In particular, we will now compute estimates for the standard errors of the income and balance logistic regression coefficients in two different ways: (1) using the bootstrap, and (2) using the standard formula for computing the standard errors in the glm() function. Do not forget to set a random seed before beginning your analysis.

Using the summary() and glm() functions, determine the estimated standard errors for the coefficients associated with income and balance in a multiple logistic regression model that uses both predictors.

```{r}
fit <- glm(default ~ income + balance, data = Default, family = "binomial")
summary(fit)
```

The standard errors obtained by bootstrapping are  
β
1
  = 5.0e-6 and  
β
2
  = 2.3e-4.
  
  

Write a function, boot.fn(), that takes as input the Default data set as well as an index of the observations, and that outputs the coefficient estimates for income and balance in the multiple logistic regression model.


```{r}
boot.fn <- function(x, i) {
  fit <- glm(default ~ income + balance, data = x[i, ], family = "binomial")
  coef(fit)[-1]
}
```



Use the boot() function together with your boot.fn() function to estimate the standard errors of the logistic regression coefficients for income and balance.

```{r}
library(boot)
set.seed(42)
boot(Default, boot.fn, R = 1000)
```

Comment on the estimated standard errors obtained using the glm() function and using your bootstrap function

The standard errors obtained by bootstrapping are similar to those estimated by glm.



In Sections 5.3.2 and 5.3.3, we saw that the cv.glm() function can be used in order to compute the LOOCV test error estimate. Alternatively, one could compute those quantities using just the glm() and predict.glm() functions, and a for loop. You will now take this approach in order to compute the LOOCV error for a simple logistic regression model on the Weekly data set. Recall that in the context of classification problems, the LOOCV error is given in (5.4).

Fit a logistic regression model that predicts Direction using Lag1 and Lag2.

```{r}
fit <- glm(Direction ~ Lag1 + Lag2, data = Weekly, family = "binomial")

```


Fit a logistic regression model that predicts Direction using Lag1 and Lag2 using all but the first observation.

```{r}
fit <- glm(Direction ~ Lag1 + Lag2, data = Weekly[-1, ], family = "binomial")

```

Use the model from (b) to predict the direction of the first observation. You can do this by predicting that the first observation will go up if  
P
(
 Direction="Up" | Lag1 , Lag2 
)
>
0.5
 . Was this observation correctly classified?
 
 
```{r}
predict(fit, newdata = Weekly[1, , drop = FALSE], type = "response") > 0.5

```
 
Yes the observation was correctly classified.

Write a for loop from  
i
=
1
  to  
i
=
n
 , where  
n
  is the number of observations in the data set, that performs each of the following steps:
Fit a logistic regression model using all but the  
i
 th observation to predict Direction using Lag1 and Lag2 .
Compute the posterior probability of the market moving up for the  
i
 th observation.
Use the posterior probability for the  
i
 th observation in order to predict whether or not the market moves up.
Determine whether or not an error was made in predicting the direction for the  
i
 th observation. If an error was made, then indicate this as a 1, and otherwise indicate it as a 0.
 
```{r}
error <- numeric(nrow(Weekly))
for (i in 1:nrow(Weekly)) {
  fit <- glm(Direction ~ Lag1 + Lag2, data = Weekly[-i, ], family = "binomial")
  p <- predict(fit, newdata = Weekly[i, , drop = FALSE], type = "response") > 0.5
  error[i] <- ifelse(p, "Down", "Up") == Weekly$Direction[i]
}
```
 
Take the average of the  
n
  numbers obtained in (d) in order to obtain the LOOCV estimate for the test error. Comment on the results.
  
```{r}
mean(error)

```
  
The LOOCV test error rate is 45% which implies that our predictions are marginally more often correct than not.

We will now perform cross-validation on a simulated data set.

Generate a simulated data set as follows:

```{r}
set.seed(1)
x <- rnorm(100)
y <- x - 2 *x^2 + rnorm(100)
```

n this data set, what is  
n
  and what is  
p
 ? Write out the model used to generate the data in equation form
 
n
  is 100 and  
p
  is 1 (there are 100 observations and  
y
  is predicted with a single variable  
x
 ). The model equation is:
y
=
−
2
x
2
+
x
+
ϵ

Create a scatterplot of  
X
  against  
Y
 . Comment on what you find.
 
```{r}
plot(x, y)

```
 
y has a (negative) quadratic relationship with  x


Set a random seed, and then compute the LOOCV errors that result from fitting the following four models using least squares:
Y
=
β
0
+
β
1
X
+
ϵ
 
Y
=
β
0
+
β
1
X
+
β
2
X
2
+
ϵ
 
Y
=
β
0
+
β
1
X
+
β
2
X
2
+
β
3
X
3
+
ϵ
 
Y
=
β
0
+
β
1
X
+
β
2
X
2
+
β
3
X
3
+
β
4
X
4
+
ϵ
 .
Note you may find it helpful to use the data.frame() function to create a single data set containing both  
X
  and  
Y
 .
 
```{r}
library(boot)
set.seed(42)
dat <- data.frame(x, y)
sapply(1:4, function(i) cv.glm(dat, glm(y ~ poly(x, i)))$delta[1])
```

Repeat (c) using another random seed, and report your results. Are your results the same as what you got in (c)? Why?

```{r}
set.seed(43)
dat <- data.frame(x, y)
sapply(1:4, function(i) cv.glm(dat, glm(y ~ poly(x, i)))$delta[1])
```

The results are the same because we are using LOOCV. When doing this, the model is fit leaving each one of the observations out in turn, and thus there is no stochasticity involved.


Which of the models in (c) had the smallest LOOCV error? Is this what you expected? Explain your answer.

The second model had the smallest LOOCV. This what would be expected since the model to generate the data was quadratic and we are measuring the test (rather than training) error rate to evaluate performance.


Comment on the statistical significance of the coefficient estimates that results from fitting each of the models in (c) using least squares. Do these results agree with the conclusions drawn based on the cross-validation results?

```{r}
for (i in 1:4) printCoefmat(coef(summary(glm(y ~ poly(x, i), data = dat))))

```

We can see that the coefficients in the first model are not highly significant, but all terms ( 
β
0
,
β
1
  and  
β
2
 ) are in the quadratic model. After this, subsequent  
β
n
  terms are not significant. Therefore, these results agree with those from cross-validation.
  

We will now consider the Boston housing data set, from the ISLR2 library.

Based on this data set, provide an estimate for the population mean of medv. Call this estimate  
^
μ
 .
 
```{r}
(mu <- mean(Boston$medv))
```


Provide an estimate of the standard error of  
^
μ
 . Interpret this result.Hint: We can compute the standard error of the sample mean by dividing the sample standard deviation by the square root of the number of observations.
 
 
```{r}
sd(Boston$medv) / sqrt(length(Boston$medv))

```
 

Now estimate the standard error of  
^
μ
  using the bootstrap. How does this compare to your answer from (b)?
  
```{r}
set.seed(42)
(bs <- boot(Boston$medv, function(v, i) mean(v[i]), 10000))
```

The standard error using the bootstrap (0.403) is very close to that obtained from the formula above (0.409).

Based on your bootstrap estimate from (c), provide a 95% confidence interval for the mean of medv. Compare it to the results obtained using t.test(Boston$medv).

Hint: You can approximate a 95% confidence interval using the formula  
[
^
μ
−
2
S
E
(
^
μ
)
,
^
μ
+
2
S
E
(
^
μ
)
]


```{r}
se <- sd(bs$t)
c(mu - 2 * se, mu + 2 * se)
```

Based on this data set, provide an estimate,  
^
μ
m
e
d
 , for the median value of medv in the population.
 
```{r}
median(Boston$medv)
```
 
We now would like to estimate the standard error of  
^
μ
m
e
d
 . Unfortunately, there is no simple formula for computing the standard error of the median. Instead, estimate the standard error of the median using the bootstrap. Comment on your findings.
 
```{r}
set.seed(42)
boot(Boston$medv, function(v, i) median(v[i]), 10000)
```
 
The estimated standard error of the median is 0.374. This is lower than the standard error of the mean.

Based on this data set, provide an estimate for the tenth percentile of medv in Boston census tracts. Call this quantity  
^
μ
0.1
 . (You can use the quantile() function.)
 
```{r}
quantile(Boston$medv, 0.1)
```

Use the bootstrap to estimate the standard error of  
^
μ
0.1
 . Comment on your findings.
 
```{r}
set.seed(42)
boot(Boston$medv, function(v, i) quantile(v[i], 0.1), 10000)
```

We get a standard error of ~0.5. This is higher than the standard error of the median. Nevertheless the standard error is quite small, thus we can be fairly confidence about the value of the 10th percentile.
 
 