# (PART\*) Section 3 {-}

# Overview {-}

<div style="color: #6D009D; font-size: 24px; font-weight: bold;text-align: justify;">
 Section 3: Manipulating and Tidying Data
</div>

::: ilos
**Learning Outcomes:**

-   ADD ILOS.
:::

**In this section, you will practice how to use the following functions:**

|   Function    |               Description               | Package |
|:-------------:|:---------------------------------------:|:-------:|
|     `c()`     |          concatenate function           | base R  |

::: attention
Remember to continue working in the RStudio project you created for this course and typing your answers in separate R scripts.
:::

# 1. Getting Started {-}

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```

To complete the practical exercises in this section, you will use data from the <a href="https://www.europeansocialsurvey.org/" target="_blank">European Social Survey (ESS) </a>. The European Social Survey is a cross-national survey that has been conducted every two years since 2002. It is one of the largest and methodologically rigorous in Europe and covers a wide range of social science topics that can be used for a wide variety of statistical analyses.  
 
::: attention
Before you begin the practical, you **MUST** create an account with ESS. Navigate to the href=" https://ess.sikt.no/en/?tab=overview" target="_blank">data portal page </a>, and click on **Log in**. The new page to which you will be taken may display text in Norwegian; switch to English using the top right button on the page. Then click on the link next to *Don't have an account on any of these providers?* to create a new account using your University of Manchester email. 
:::

For this practical, you will perform tasks using an earlier version of data from **ESS Round 9 (2018)**. There are two files that you will be using in this practical:   

**ESS9e01_1.dta** (individual-level data)   
**ESSMD-2018-cntry_F1.dta** (country-level data).   

Both of these data files can be downloaded from the learning materials pages on Blackboard.

::: folder
Ensure that you download and place these two files, as usual, in your `data` folder of your RStudio project for this course. 
::: 

# 2. Data Import {-}

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


## 2.1 Task {-}

Remember the first step before beginning any analysis in R?  

You will need to load two packages for this session, `tidyverse` and `haven`.

```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(haven)
```

## 2.2 Task {-}

Import the two data files. Name  `ESS9e01_1.dta` as **ess9** and `ESSMD-2018-cntry_F1.dta` as **ess_country_full**. 

```{r warning = FALSE, message = FALSE}
ess9 <- read_dta("data/ESS9e01_1.dta")

ess_country_full <- read_dta("data/ESSMD-2018-cntry_F1.dta")
```

## 2.3 Task {-}

Use the `glimpse()` function from `dplyr` to explore the structure of **ess9**. 

```{r, results = FALSE}
glimpse(ess9)
```

::: question
How does the output of `glimpse()` differ from the output of the `str()` function?
:::

::: {.r-comment style="display: none;"}
As you can see, `glimpse()` provides a similar output to the base R `str()` function but it is adapted specifically for tibble data objects. Since this object **ess9** is very large, the output is rather difficult to navigate.
:::

# 3. Subsetting {-}


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


In general, datasets will require some degree of subsetting of variables that would be most relevant to your topic of analysis. Normally, you would study the data dictionary that provides details about each variable, what it represent, and how it is coded. For this practical, we will use the following variables: 

## 3.1 Task {-}

Subset **ess9** so that it contains only the following variables: idno, cntry, imbgeco, imueclt, imwbcnt, vote, uempla, agea, eduyrs, gndr, region.  

Name this new data object **ess**.  

```{r}

# Without pipe 
ess <- select(ess9, idno, cntry, imbgeco, imueclt, imwbcnt, vote, 
              uempla, agea, eduyrs, gndr, region)

# With pipe 
ess <- ess9 %>% select(idno, cntry, imbgeco, imueclt, imwbcnt, vote, 
                       uempla, agea, eduyrs, gndr, region)
```

::: {.r-comment style="display: none;"}
In general, datasets will require some degree of subsetting of variables that would be most relevant to your topic of analysis. Normally, you would study the data dictionary that provides details about each variable, what it represent, and how it is coded. For this practical, we will only need the variables you have just subsetted.
:::

## 3.2 Task {-}

Explore **ess** using `glimpse()`.

```{r}
glimpse(ess)
```

::: question
What dimensions does the object have? What do you think about the class of some of these variables; do you think any of these might require transformation?
:::

::: {.r-comment style="display: none;"}
This object is quite large (over 30,000 observations for 11 variables). You can also see that at least two variables, such as **gndr** and **cntry** should be coded as factors; in the dataset, they are coded as numeric or character, respectively.
:::

# 4. Filtering Observations {-}


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


## 4.1 Task {-}

Explore the **edurys** variable and find out what this variable measures. 

```{r}
attributes(ess$eduyrs)
```

Count the number of observations. Are there any missing values? 

```{r}
count(ess, eduyrs)
```

Use `filter()` remove the missing values from the variable. 

```{r}
ess <- filter(ess, !is.na(eduyrs))
```

Confirm that the missing values have been removed. 

```{r}
sum(is.na(ess$eduyrs))
```

## 4.2 Task {-}

Explore the **vote** variable and find out what this variable measures. 

```{r}
attributes(ess$vote)
```

::: {.r-comment style="display: none;"}
The **vote** variable records whether survey respondents voted during their last national election and is a categorical variable with 3 categories (yes, no, not eligible) coded as a numeric variable.
:::

Count the number of observations. Are there any missing values? 

```{r}
count(ess, vote)
```

::: {.r-comment style="display: none;"}
The **vote** variable highlights that (by far), the majority of the respondents who participated in the survey did vote in their last national election. There are a total of 412 missing values.
:::

Count the number of observations but this time, drop the missing values. 

```{r}
ess %>% count(vote) %>%
        drop_na()
```

::: {.r-comment style="display: none;"}
To count the values of the vote variable but drop the missing ones, we can use the `count()` and `drop_na()` functions in the same sequence by connecting these two functions with the pipe operator.  

Another way to obtain the same result is to first remove the missing values by using the `filter()` function and then use `count()`.     

`ess %>% filter(!is.na(vote)) %>% count(vote)`
:::


## 4.3 Task {-}

Explore the **gndr** and find out what it measures.

```{r}
attributes(ess$gndr)
```

::: {.r-comment style="display: none;"}
The **gndr** variable records whether the sex of the surey respondents is a categorical variable with 2 categories (male, female) also coded as a numeric variable.
:::

Find the number of cases for each category. Are these are missing values?

```{r}
count(ess, gndr)
```


# 5. Keys and Unique Values {-}

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


To perform the remaining exercises, you will need to join the individual-level data object **ess** with the country-level data object **ess_country_full**.

## 5.1 Task {-}

Explore **ess_country_full** using `glimpse()`. 

```{r}
glimpse(ess_country_full)
```

:::question 
Do you notice any patterns in the names of the variables?
:::

::: {.r-comment style="display: none;"}
When using `glimpse()`, it is evidently cumbersome to explore the variables in the console. Nevertheless, if you glance through the output in the console, you can note several patterns in the names. For example, each variable starts with the `c_` pattern, and each variable name ends with **year**. Variables contain specific name patterns that show which variables were recorded in multiple years. 
:::

## 5.2 Task {-}

The data file contains country level data from multiple years. However, since our individual level data only contains data from 2018, the **ess_country_full** data object must be subsetted so that it also contains data only from 2018. 

Subset **ess_country_full** such that it only contains the variable **cntry** and variables from 2018. 

```{r}
ess_cntry18 <- ess_country_full %>% select(cntry, ends_with("2018"))
```

Now confirm that the changes have been applied. 

```{r}
glimpse(ess_cntry18)
```

::: {.r-comment style="display: none;"}
When using `glimpse()`, it is evidently cumbersome to explore the variables in the console. Nevertheless, if you glance through the output in the console, you can note several patterns in the names. For example, each variable starts with the `c_` pattern, and each variable name ends with **year**. Variables contain specific name patterns that show which variables were recorded in multiple years. 
:::

## 5.3 Task {-}

Before merging, we must check whether the two data objects (**ess** and  **ess_cntry18**) have unique keys. One sensible option for unique key would be the **cntry** variable which is found in both objects. 

Check whether the observations for the **cntry** variable in **ess_cntry18** are unique. 

```{r}
ess_cntry18 %>% 
group_by(cntry) %>% 
count() %>% 
filter(n > 1) 
```

::: {.r-comment style="display: none;"}
For **ess_cntry18**, the **cntry** variable has unique observations and therefore, can be used as a key.
:::

Now, do the same for **ess**. 

```{r}
ess %>% 
group_by(cntry) %>% 
count() %>% 
filter(n > 1) 
```

::: {.r-comment style="display: none;"}
In **ess**, the **cntry** variable has multiple observations, and therefore, does **not* represent a unique key. 
:::

::: question 
If **cntry** does not represent a unique key, can **idno** be used instead?  
:::


```{r}
ess %>% 
group_by(idno) %>% 
count() %>% 
filter(n > 1) 
```

::: {.r-comment style="display: none;"}
In **ess**, the **idno** variable has multiple observations, and therefore, does not represent a unique key either. 
:::

## 5.4 Task {-}

If neither **cntry** nor **idno** are unique keys for **ess**, then create a unique key called **id**, place this variable at the start of the dataset, and arrange the values in ascending order.

```{r}
ess <- ess %>%
arrange(cntry, idno) %>% 
mutate(id = row_number())
```

Don't forget to double-check if the changes you have made have been applied. 

```{r}
ess %>% 
group_by(id) %>% 
count() %>% 
filter(n > 1)
```

## 5.5. Task {-}

In the context of this practical, we are interested to explore individual level differences given country. Therefore, the purpose of the merge is to add the country level data to the individual level data so that each survey respondent will be associated with a particular country.  

:::question 
What kind of join will this be?
:::

::: {.r-comment style="display: none;"}
This will be a `left_join()` with the individual level data as the first input and the **cntry** variable as the key since this is the key in the country level data object. 
:::

For information purposes, it is important to explore whether all countries are available in the individual level data object so that we can identify whether any information would be lost during the merge. 

To do so, we can use the `unique()` base R function which identifies unique values in variables and the  `%in%` logical operator which is used to compare values and produce a `TRUE` or `FALSE` output. 

To check for unique values in the **ess** object, you can run the code below in your console. 

`unique(ess$cntry)`  

We should also find out if all the countries in the country level data are also found in the individual level data. Essentially, the below line of code is asking R whether the values of the variable `cntry` in the `ess` object are are also present in the `ess_cntry18` object. Run the code below in your console.

`country_present <- unique(ess$cntry) %in% unique(ess_cntry18$cntry)`  

If you print the contents, you will find that there are multiple `FALSE` instances, which means that there are countries that are not present in the country level dataset.

To find out exactly which countries are not present in the country level data, run the below code in your console. This combines the unique function to check for countries in the `ess` object and the subsetting square brackets so that the output excludes countries that are present in both objects and provides only those which are absent from `ess_cntry18`. Run the code below in your console.

`unique(ess$cntry)[!country_present]`


# 6. Perfoming the Merge {-}


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


## 6.1 Task {-}

Merge **ess** with **ess_cntry18** by **cntry** and name this new data object **ess2**.

```{r}
ess2 <- left_join(ess, ess_cntry18, by = "cntry")
```

## 6.2 Task {-}

Explore the object using `glimpse()`. What are the dimensions of this new data object? 

```{r}
glimpse(ess2)
```

::: {.r-comment style="display: none;"}
This new object has 35,510 observations of 21 variables.
:::

# 7. Data Transformation {-}


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```


Now that our new dataset contains all the information needed, let's have a look at the variables and perform some transformations. To find out more about what the variable code names mean, you can of course consult the data dictionaries for more detailed information, but you can also use the `attributes()` function. 

## 7.1 Task {-}

Explore the **gndr** variable using the `count()` function. 

## 7.2 Task {-}

```{r}
count(ess2, gndr)
```

Transform the **gndr** variable into a factor variable called **sex** with two categories; assign a Male label to values of 1 and a Female label to the values of 2. Use the `count` function to confirm the changes.  

```{r}
ess2$sex <- factor(ess2$gndr, labels = c("Male", "Female"))
```

Use `count()` to confirm the changes.  

```{r}
count(ess2, sex)
```


## 7.3 Task {-}

Find out what the **imbgeco** variable measures. 

```{r}
attributes(ess2$imbgeco)
```

Transform the **imbgeco** variable into a factor variable called **immigration** with three categories as follows: 

- values 0 to 3 should represent "Bad for Economy"  
- values 4 to 6 should represent "Neutral"  
- values 7 to 10 should represent "Good for Economy"  

```{r}
ess2 <- mutate(ess2, 
                   immigration = case_when(
                     imbgeco >=0 & imbgeco <= 3 ~ "Bad for Economy",
                     imbgeco >= 4 & imbgeco <= 6 ~ "Neutral",
                     imbgeco >= 7 & imbgeco <= 10 ~ "Good for Economy"), 
                   immigration_factor = factor(immigration))
```


Confirm the changes using the `count()` function. 

```{r}
count(ess2, immigration_factor)
```

## 7.4 Task {-}

Create a new factor variable called **immigrants** based on the **imwbcnt** variable such that:  
- values of 0 to 3 represent "Worse"  
- values of 4 to 6 represent "Neither Better nor Worse"  
- values of 7 to 10 represent "Better"  

```{r}
ess2 <- mutate(ess2, 
                   immigrants = case_when(
                     imbgeco >=0 & imbgeco <= 3 ~ "Worse",
                     imbgeco >= 4 & imbgeco <= 6 ~ "Neither Better nor Worse",
                     imbgeco >= 7 & imbgeco <= 10 ~ "Better"), 
                   immigrats = factor(immigrants))
```

Confirm the changes using the `count()` function. 

```{r}
count(ess2, immigrants)
```


Use the appropriate `forcats` function to assign NA values to a level called "missing" for the **immigrants** variable.

```{r}
ess2 <- mutate(ess2, 
	             immigrants = fct_explicit_na(immigrants, 
					                                  na_level = "Missing"))
```

Confirm the changes. 

```{r}
count(ess2, immigrants)
```

## 7.5 Task {-}

Remove the "c_" prefix and the "_2018" suffix from the **ess2** variables. Now use the appropriate function to check whether the changes have been applied. 

```{r}

# Removing the "c_prefix"
ess2 <- ess2 %>%
rename_all(~ str_remove(., "c_"))

# Removing the "_2018" suffix
ess2 <- ess2 %>%
rename_all(~ str_remove(., "_2018"))
```

Now use a suitable function to check whether the changes have been applied. 

```{r}
names(ess2)
```


# 8. Saving Data Objects {-}


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```

Now export the **ess2** data object as a `.csv` file using the `write_csv()` function from the `readr` core tidyverse package. Name this file **ess2**. 

::: folder 
Make sure that you export this file in your **data** subfolder in your RStudio project working directory. You will use this file in other practicals of this course. 
:::

```{r, eval = FALSE}
write_csv(ess2, "data/ess2.csv")
```

:::question 
Why would we use `write_csv()` instead of `write.csv()`?
:::

::: {.r-comment style="display: none;"}
The `write_csv()` function works in a similar way to the base R function `write.csv()` with which you are already familiar. The main difference between the two is that the `tidyverse` function maintains the metadata associated with the data object whilst the base R function does not. It is therefore more useful when you re-import the data, particularly if you have not completed the data manipulation stages.  
:::

