---
editor_options: 
  markdown: 
    wrap: 72
---

# Practical 2: Foreign Direct Investment (FDI) Study {-}

*This practical has been developed by Dr. Tatjana Kecojevic, Lecturer in Social Statistics.*


::: file
For the tasks below, you will require the **FDI** dataset.  

Click here to download the file:
<a href="data/FDI.csv" download="FDI.csv"> FDI.csv </a>.

Remember to place your data file in a separate subfolder within your R
project working directory.
:::

A business consultancy firm is compiling a major report about
globalisation. One aspect of this study concerns the determinants of FDI
undertaken by multi-national enterprises. Relevant information from a
sample of 60 multi-national companies that undertook significant
investment in overseas projects was made available as follows:

+---------------+----------------------------------------------------+
| Variable Name | Variable Description                               |
+:=============:+:==================================================:+
| FDI           | Value of FDI undertaken, in £ millions, by          |
|               | investing company                                  |
+---------------+----------------------------------------------------+
| GDP_Cap       | GDP per capita, £000s, in the country receiving    |
|               | the investment                                     |
+---------------+----------------------------------------------------+
| Gr_rate       | The economic growth rate, in %-terms, in the       |
|               | country receiving the investment                   |
+---------------+----------------------------------------------------+
| ROC           | The average return on capital invested, in         |
|               | %-terms, in the country receiving the investment   |
+---------------+----------------------------------------------------+
| Stable        | The political stability of the country receiving   |
|               | the investment as measured by the number of        |
|               | changes in government over the past 25 years       |
+---------------+----------------------------------------------------+
| Infra         | Infrastructure facilities (eg transport,           |
|               | communications) in the country receiving the       |
|               | investment                                         |
|               |                                                    |
|               | Coded: 1 = basic infrastructure 2 = good           |
|               | infrastructure                                     |
+---------------+----------------------------------------------------+
| Trade         | The openness to trade of the country receiving the |
|               | investment                                         |
|               |                                                    |
|               | Coded: 1 = trade tightly controlled 2 = some       |
|               | restrictions on trade 3 = free trade               |
+---------------+----------------------------------------------------+

This is a multiple regression type of the problem; FDI is the key
response variable as this study concerns the determinants of FDI
undertaken by multi-national enterprises.

The model is defined as $Y = b_0 + b_1x_1 + b_2x_2 + ... + b_kx_k + e$,
for the general $k$ explanatory variable model and where e is also known
as the error term $e ∼ N(0,\sigma^2)$, with the error term from a normal
distribution with a mean of $0$, and a variance of $\sigma^2$.

Based on prior knowledge, we make the assumption that GDP_Cap, Gr_rate,
ROC, Infra, and Trade have positive relationships with FDI, whilst
Stable has a negative relationship with FDI.

We will use our best fit model to predict FDI for the following information:  *country X receiving the investment has GDP per capita of 11.1 and Gr_rate per capita of 3.05; The average return on capital invested is 20.5%; There were 11 changes of government over the past 25 years and country X has good infrastructure with some restrictions on trade.*

First, let's load the required packages: 

```{r warning = FALSE, message = FALSE}
library(tidyverse)
# you should have already installed this package as part of the previous Demonstration
library(GGally)
```

Let's import the data into R.

```{r}
mydataq1 <- read.csv("data/FDI.csv", header = T) 
```

Now let's get a glimpse of the data. As you know, there are many ways to
do that, such as, for example, using the tidyverse `glimpse` function.
This is quite a handy function because it also tells us more about the
class of each variable. We can see that although **Infra** and **Trade**
categorical, these are coded as integers.

```{r}
glimpse(mydataq1)
```

We therefore need to transform them into factors.

```{r}
mydataq1 <- mydataq1 %>%
  mutate(Infra = as_factor(Infra),
         Trade = as_factor(Trade)
         )
```

We can then explore all variables in the dataset as pairs using a matrix
of plots. Among many interesting features, we can note quite strong
correlations among pairs of variables which suggest the presence of
multicollinearity: GDP_Cap and ROC, Infra and GDP_Cap, Infra and ROC,
and Infra and Gr_rate.

```{r warning = FALSE, message = FALSE}
GGally::ggpairs(mydataq1)
```

Ok, so our initial model is:

$FDI = b_0 + b_1GDP\_Cap + b_2Gr\_rate + b_3ROC – b_4Stable + b_5Infra + b_6Trade + e$

where **Infra** and **Stable** are dummy variables.

We can have a look at how these two dummy variables are used in the
model by using the `contrasts()` function from base R.

The **Infra** variable is coded as *1 = basic infrastructure* and *2 =
good infrastructure*. Since this is a binary variable, there will be one
reference category and a single dummy variable.

```{r}
contrasts(mydataq1$Infra)
```

The **Trade** variable is coded as *1 = trade tightly* and *2 = some
restrictions on trade* and *2 = some restrictions on trade*. Since this
is variable with three categories, there will be one reference category
and two dummy variables.

```{r}
contrasts(mydataq1$Trade)
```

Let's now fit our multiple regression model.

```{r}
m1 <- lm(FDI ~ GDP_Cap + Gr_rate + ROC + Stable + Infra + Trade, data = mydataq1)
```

And explore the results.

```{r}
summary(m1)
```

:::question
What do the results indicate?   
How do they compare with our initial assumptions?   
How do you proceed with the analysis?
:::

