---
editor_options: 
  markdown: 
    wrap: 72
---

# Demonstration 2: Principal Component Analysis in R {-}

For the tasks below, you will require the **World Happiness Report (2021)** dataset.  

Click here to download the file: 
<a href="data/world_happiness_report_2021.csv" download="world_happiness_report_2021.csv"> world_happiness_report_2021.csv </a>.   
Remember to place your data file in a separate subfolder within your R project working directory.  

Prior to beginning the practical, you will be required install two packages `easyCODA` and `janitor`. 

## Load packages {-}

```{r, warning = FALSE, message = FALSE}
library(easyCODA)
library(dplyr)
library(tidyr)
library(readr)
library(janitor)
library(ggplot2)

theme_set(theme_light())
```

## Preprocess the world happiness report data {-}

```{r}
happy <-
  read_csv(
    file = "data/world_happiness_report_2021.csv",
    show_col_types = FALSE
  ) |>
  clean_names()

happy_standardized <- 
  happy |>
  select(
    social = social_support,
    life = healthy_life_expectancy,
    choices = freedom_to_make_life_choices,
    generosity,
    corruption = perceptions_of_corruption
  ) |>
  scale()

rownames(happy_standardized) <- happy$country_name
```

## Profile the happiness data {-}

```{r}
head(happy)
```

## Plot happiness score against feature {-}

```{r}
happy |>
  select(
    ladder = ladder_score,
    social = social_support,
    life = healthy_life_expectancy,
    choices = freedom_to_make_life_choices,
    generosity,
    corruption = perceptions_of_corruption
  ) |>
  pivot_longer(-ladder, names_to = "feature", values_to = "value") |>
  ggplot(aes(value, ladder)) +
  geom_point(alpha = 0.4) +
  scale_x_continuous("Feature value") +
  scale_y_continuous("Cantrill Ladder happiness score") +
  facet_wrap(~ feature, scales = "free_x")
```

## Run PCA using five features {-}

```{r}
happy_pca <- PCA(happy_standardized, weight = FALSE)
summary(happy_pca)
```

:::question
What is the percentage of variance explained by each principal component?
:::

```{r}
happy_pca$sv ^ 2 / sum(happy_pca$sv ^ 2)
```

:::question
What is the cumulative variance explained for each variable?
:::

```{r}
vexp <- function(feature, df, pca) {
  reg_pc1 <- lm(df[, feature] ~ pca$rowpcoord[, 1])
  reg_pc1_pc2 <- lm(df[, feature] ~ pca$rowpcoord[, 1] + pca$rowpcoord[, 2])
  
  explained_pc1 <- cor(predict(reg_pc1), df[, feature]) ^ 2
  explained_pc1_pc2 <- cor(predict(reg_pc1_pc2), df[, feature]) ^ 2
  
  c(explained_pc1, explained_pc1_pc2)
}
```

```{r}
var_explained <- sapply(1:5, \(x) vexp(x, happy_standardized, happy_pca))
colnames(var_explained) <- colnames(happy_standardized)
rownames(var_explained) <- c("PC1", "PC2")
t(round(var_explained, 3))

round(colMeans(t(var_explained)), 3)
```

:::question
How do we build a correlation matrix of all variables?
:::

```{r}
round(cor(cbind("ladder" = happy$ladder_score, happy_standardized)), 3)
```

## Correlations of happiness score with PC1 and PC2 {-}

```{r}
row_principal <- -happy_pca$rowpcoord[, 1:2]
colnames(row_principal) <- c("PC1", "PC2")
round(cor(cbind("ladder" = happy$ladder_score, row_principal)), 3)
```

## Plot first two principal components {-}

```{r}
row_principal |>
  as_tibble() |>
  mutate(
    region = happy$regional_indicator,
    country = happy$country_name
  ) |>
  ggplot(aes(PC1, -PC2, color = region, label = country)) +
  geom_hline(yintercept = 0, color = "gray70", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray70", linetype = "dashed") +
  geom_text(size = 3, show.legend = FALSE) +
  coord_equal()
```

## Regress happiness on the five indicators {-}

```{r}
happy_standardized_pc <-
  happy_standardized |>
  as_tibble() |>
  mutate(
    ladder = happy$ladder_score,
    pc1 = happy_pca$rowpcoord[, 1],
    pc2 = happy_pca$rowpcoord[, 2],
    country = happy$country_name,
    region = happy$regional_indicator
  )

happiness_full <- 
  lm(ladder ~ social + life + choices + generosity + corruption,
     data = happy_standardized_pc)

summary(happiness_full)
```

:::question
What is the correlation between observed y and fitted y?
:::

```{r}
sqrt(summary(happiness_full)$r.squared)
```

## Visualise happiness against principal components {-}

```{r}
happy_standardized_pc |>
  pivot_longer(
    cols = c(pc1, pc2),
    names_to = "PC",
    values_to = "value"
  ) |>
  ggplot(aes(value, ladder)) +
  geom_point(aes(color = region)) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    linewidth = 0.5,
    color = "grey20",
    se = FALSE
  ) +
  scale_x_continuous("PC Value") +
  scale_y_continuous("Cantrill Ladder Happiness Score") +
  facet_wrap(~ PC) +
  theme(text = element_text(size = 12))
```

## Regress happiness on first principal component {-}

```{r}
happiness_reduced <-
  lm(formula = ladder ~ pc1, data = happy_standardized_pc)

summary(happiness_reduced)
sqrt(summary(happiness_reduced)$r.squared)
```

## 👉 TASK {-}

Now let's consider PCA using six features. In the example of PCA above, we did not use `logged_gdp_per_capita` feature for the PCA analysis.  

Run PCA using the original five features (`ladder`, `social`, `life`, 
`choices`, `corruption`) and `logged_gdp_per_capita`. You may need to do 
some preprocessing first.  

> Your code here  

Now regress happiness on the first principal component from your analysis.  

> Your code here  

How does this compare to the previous analysis with five features? Consider 
here the R^2 value and the correlation between the observed and fitted values.  

> Your code here  

