# (PART\*) Section 7 {-}

# Overview {-}

<div style="color: #6D009D; font-size: 24px; font-weight: bold;text-align: justify;">
 Section 7: Programming with R
</div>

::: ilos
**Learning Outcomes:**

-   ADD ILOS.
:::

**In this section, you will practice how to use the following functions:**

|   Function    |               Description               | Package |
|:-------------:|:---------------------------------------:|:-------:|
|     `c()`     |          concatenate function           | base R  |

::: attention
**WARNING 1:** Before proceeding to the exercises, ensure that you create and work in a RStudio project and that you **write** and **save** your answers to the exercises in an R script.

**WARNING 2:** To adequately develop your practical programming skills, it is **essential** that you attempt all exercises by typing out the code rather than simply copy-pasting the answers.
:::

You will require the `tidyverse` package for this practical. 

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
```

# 1. Basic Functions and Loops {-}

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```

## 1.1 Task {-}

Write a function that calculates the sum between two values and name it `sum_func`. Use values 55 and 43 to test your function.

```{r}
sum_func <- function(a, b){
  c <- a + b
  return(c)
}
```

## 1.2 Task {-}

Use values 55 and 43 to test your function.

```{r}
sum_func(55,43)
```

## 1.3 Task {-}

Write a function that calculates the average (mean) of two numbers and name it `av_func`. *Hint: divide the sum of the values by the length of the vector.*  

```{r}
av_func <- function(x){
  avg <- sum(x)/length(x)
  return(avg)
}
```

Then, use a loop to apply this function to the first 3 columns of the **mtcars** `tidyverse` built-in dataset and store it in a vector of type double called **mean_mtcars**. Print the results

```{r}
mean_mtcars <- vector("double", length = 3)
  for(i in seq_along(mtcars[1:3])) {
    mean_mtcars[[i]] <- av_func(mtcars[[i]])
  }

mean_mtcars
```

## 1.4 Task {-}

Write a loop that prints values to the power of 2 in a sequence from 1 to 6. 

```{r}
for(i in 1:6) {
   print(i^2)
}
```

## 1.5 Task {-}

You have an object x which equals 205. Write a function with the following three conditions and store it in an object called **example**:   

- if x is less than 10, then return "x is smaller than 10"   
- if x greater than 10, then return "x is larger than 10"    
- if x is equal to 10, then return "x is exactly 10"   
 
Print the contents of the object. 


```{r}
x <- 205
example <- if (x < 10) {
    "x is smaller than 10"
  } else if (x > 10) {
    "x is larger than 10"
  } else if (x == 10) {
    "x is exactly 10"
  }

print(example)
```

# 2. Summary Statistics and Plots

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre > code');

    codeBlocks.forEach(function(block) {
        // Check if the block's content starts with '##'
        if (block.textContent.trim().startsWith('##')) {
            var button = document.createElement('button');
            button.className = 'toggle-button';
            button.textContent = 'Show R Output';
            button.style.display = 'block'; // Ensure button is visible
            block.style.display = 'none'; // Initially hide the R output

            button.onclick = function() {
                if (block.style.display === 'none') {
                    block.style.display = 'block';
                    button.textContent = 'Hide R Output';
                } else {
                    block.style.display = 'none';
                    button.textContent = 'Show R Output';
                }
            };

            block.parentNode.insertBefore(button, block);
        }
    });
});
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var rComments = document.querySelectorAll('.r-comment');

    rComments.forEach(function(comment) {
        // Create the toggle button
        var button = document.createElement('button');
        button.className = 'toggle-comment-button';
        button.textContent = 'Show Explanation';
        button.style.display = 'block'; // Ensure button is visible
        comment.style.display = 'none'; // Initially hide the comment

        // Add click event listener to the button
        button.onclick = function() {
            if (comment.style.display === 'none') {
                comment.style.display = 'block'; // Adjust as needed
                button.textContent = 'Hide Explanation';
            } else {
                comment.style.display = 'none';
                button.textContent = 'Show Explanation';
            }
        };

        // Insert the button before the comment
        comment.parentNode.insertBefore(button, comment);
    });
});

```

## 2.1 Task {-}

Create a loop to calculate the standard deviation of each column of the **mtcars** dataset. Print the output

```{r}
output <- vector("double", ncol(mtcars))
  for (i in seq_along(mtcars)) {
    output[[i]] <- sd(mtcars[[i]])
  }

output
```

## 2.2 Task {-}

Use an appropriate base R functional to calculate the standard deviation of each column of the **mtcars** dataset and returns a vector.

```{r}
sapply(mtcars, sd)
```

## 2.3 Task {-}

Use an appropriate `tidyverse` functional to calculate the standard deviation of each column of the **mtcars** dataset and returns a vector of type `double`.

```{r}
map_dbl(mtcars, sd)
```

## 2.4 Task {-}

Using an appropriate base R functional, calculate the median age by religion in the **gss_cat** built-in dataset from `tidyverse`. 

```{r}
tapply(gss_cat$age, gss_cat$relig, median, na.rm = T)
```

## 2.5 Task {-}

Using an appropriate base R functional, calculate the mean tv hours by party ID in the **gss_cat** dataset. 

```{r}
tapply(gss_cat$tvhours, gss_cat$partyid, mean, na.rm = T)
```

## 2.6 Task {-}

Write a function that produces plots of variables **rincome**, **relig**, and **tvhours** for the **gss_cat** dataset. Specify these variables by name rather than by index.

```{r}
for(i in gss_cat[c("rincome", "relig", "tvhours")]){
  p <- ggplot(gss_cat, aes(x = i)) +
        geom_bar()
  print(p)
}
```

## 2.7 Task {-}

Create a function called `make_center` that takes as an input a vector and centers it (i.e. subtract the mean from x). Add a check for the input to be numeric or otherwise return an error message that states "x not numeric".

```{r}
make_center <- function(x) {
  if(!is.numeric(x)) {
    stop("x not numeric")
  }
  x - mean(x, na.rm = T)
}
```

## 2.8 Task {-}

Apply the `make_center()` function you just created to center all variables of the **mtcars** dataset using a functional that returns a data frame as output.

```{r}
map_df(mtcars, make_center)
```

## 2.9 Task {-}

Apply the `make_center()` function to center all variables of the **mtcars** dataset using a loop that stores the output as a list object; name this list object **center_list** and print the output.

```{r}
center_list <- list(NULL)
  for (i in seq_along(mtcars)) {
    center_list[[i]] <- make_center(mtcars[[i]]) 
  }

center_list
```

