# Introducing Statistical Programming with R and RStudio

**Learning Outcomes:** 

*- Employ the basics of the R language to run simple lines of code;*  
*- Import and export data objects.* 

## Creating and Exploring Vector Objects

### Exercise 


```{css, echo=FALSE}
/* Button styles */
.toggle-button {
    display: block;
    margin: 10px 0;
    padding: 5px 10px;
    background-color: #007bff;
    color: white;
    border: none;
    border-radius: 5px;
    cursor: pointer;
}

/* Style for the <pre><code> blocks */
pre {
    background-color: #f8f8f8; /* Light grey background */
    border: 1px solid #ddd; /* Light grey border */
    padding: 10px; /* Padding around the text */
    border-radius: 4px; /* Rounded corners */
    overflow: auto; /* Add a scrollbar if the content is too wide */
}

code {
    font-family: 'Courier New', monospace; /* Monospace font for code */
    color: #333; /* Dark grey text color */
}

/* Initially hide code blocks */
pre {
    display: none;
}
```


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
    var codeBlocks = document.querySelectorAll('pre.sourceCode');

    codeBlocks.forEach(function(block, index) {
        var button = document.createElement('button');
        button.className = 'toggle-button';
        button.textContent = 'Show Code';
        button.addEventListener('click', function() {
            if (block.style.display === 'none') {
                block.style.display = 'block';
                button.textContent = 'Hide Code';
            } else {
                block.style.display = 'none';
                button.textContent = 'Show Code';
            }
        });

        block.before(button);
    });
});
```

Create a vector object called `num_vct` that contains the following values: 70, 8, 50, 100.

```{r}
num_vct <- c(70, 8, 50, 100) 
```

Print the contents. 

```{r}
num_vct 
```

a. Use the class() function to find out what type of object `num_vct` is.
b. Multiply this vector by 2.
c. Divide the vector by 10. 
d. Subtract 5 from the vector. 
e. Add 8 to the vector.

### Exercise

Create a character vector object that contains the following: I, love, programming, with, R.  Name this vector `char_vct` and print the contents.

a. Use the class() function to find out what type of object `char_vct` is.
b. Perform tasks b to e from Exercise 1. What results do you obtain? Why?

### Exercise

Create a new numeric vector that contains the following values: 5, 7, 9, 10. Name it `num_vct2` and print the contents. 

a. Add num_vct2 to num_vct. 
b. Divide num_vct by num_vct2. 
c. Multiply num_vct2 by num_vct. 
d. Subtract num_vct from num_vct2

### Exercise

The number of elements in a vector is referred to as the length of the vector. With small vectors such as those you have already created, you can simply count these. Alternatively, the `length()` function can be used. 

Use the length function to obtain the length of the following objects: `num_vct`, `char_vct`, `num_vct2`.

### Exercise

Create a vector object that contains the following logical and numeric values: TRUE, 6, FALSE, 10, FALSE. Name it `new_vct` and print the output. 

Does the output match the contents you entered when you created the vector? Why?

## Importing and Exporting Data

Throughout the course units of this programme, you will utilise real datasets to develop your data analysis and interpretation skills. 

For this exercise, you will navigate to the UK Data Service (UKDS) website and acquire the British Social Attitudes Survey (2019), Poverty and Welfare: Open Access Teaching Dataset. 

### Exercise 

Navigate to the British Social Attitudes Survey (2019) webpage on the UK Data Service website (https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=8850#!/access-data)

Scroll down to the end of the webpage and download the dataset in both SPSS and STATA formats in your R project working directory.

To access the SPSS data file, unzip folder and open  the folders in the following sequence: UKDA-8850-spss > spss > spss25. Follow the same approach for accessing the STATA file. Place these two files in your root R project working directory or within a sub-folder in this directory. 

a. Import the dataset in SPSS format; name this object `bsas_spss`. 
b. Import the dataset in STATA format; name this object `bsas_stata`. 

### Exercise

a. How many observations and variables do the `bsas_spss` and `bsas_stata` objects have? Are these identical for both objects?
b. Use the view function to explore these data objects. 

### Exercise

Export both `bsas_spss` and `bsas_stata` data objects in `.RData` format. 

### Exercise

Export `bsas_spss` as a single object in `.RDS` format.  

### Exercise

Export the `bsas_stata` object as a `.csv` file. 

### Exercise

a. Import the `.csv` file you have just created and name this object `bsas_stata2`. View the contents of the `bsas_stata2` and `bsas_stata` objects. What are the differences between the two, if any? Why?
b. Import the spss you have created and name this object `bsas_spss2`. 
c. Load the `.Rdata` file your created. 

## Bonus Task: Data on GitHub

**Importing Data from Github Repositories**

Many different types of data can be imported in R using either base R functions or functions from packages. However, R is not limited to importing 'hard copy' files from your machine but also supports direct import of data files located on websites for example. 

A Github Repository is a cloud-based, online platform that allows programming users to openly share research projects, associated documentation, data files, and comments with other users and/or the public. 
Explore the Johns Hopkins Whiting School of Engineering COVID-19 Github Repository (https://github.com/CSSEGISandData/COVID-19) 

1. Open the csse_covid_19_data folder found at the beginning of the page. 
2. Then open the csse_covid_19_time_series folder. 
3. Click on time_series_covid19_confirmed_global.csv link. 
4. Click on View Raw. 
5. The webpage will now show you a long series of numbers. This is the 'data file' that you need to import. Note that this file is a `.csv` file. The same base R function you have previously used will also work in this case. 

To import the data file, create a new object called `covid_johnshop`. Within the base R `.csv` function, paste the full data file web link. Do not forget to enclose this link with quotation marks. 

View the contents of this object. How many observations and variables does this object have? What class is this object?





